The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-05 rational.tex
intro a bit

In a recent paper~\cite{infeffects}, we argued that, by encompassing
irreversible physical primitives, conventional abstract models of computation
have also inadvertently included some \emph{implicit} computational effects
which we called \emph{information effects}. We then developed a pure
reversible model of computation that is obtained from the type isomorphisms
and categorical structures that underlie models of linear logic and quantum
computing and that treats information as a linear resource that can neither
be erased nor duplicated. In this paper, we show that our pure reversible
model unveils deeper and more elegant symmetries of computation than have
previously been reported. In particular, we expose two notions of duality of
computation: an additive duality and a multiplicative duality that give rise
to negative types and fractional types respectively. Although these types
have previously appeared in the literature (see Sec.~\ref{sec:related}), they
have typically appeared in the context of conventional languages with
information effects, which obscured their appeal and murked their properties.

@@

Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism with a non-trivial and interesting computational
interpretation.  On the left we have a producer that can supply a value of
type $a$ and on the right we have a consumer that demands a value of type
$a$. For the purposes of the example, think of the producer as someone having
\$20.00 and the consumer as someone demanding \$20.00. It is certainly
possible to directly connect the producer and consumer: such a connection
would be modeled by the identity function. The proof above implements this
connection in a much more interesting way, however. A transformation $0 =
(-a) + a$ is used at the consumer site to create money and a corresponding
debt. The consumer gets the money immediately and the debt travels
``backwards'' towards the producer where it is paid.  In other words, this
computation models something similar to a credit card transaction. We discuss
this example in detail in Sec.~\ref{sub:neg} and only focus on two important
points here:
\begin{itemize}
\item As the example illustrates, negative types corresponds to ``debts.''
  It would clearly be disastrous if debts could be deleted or
  duplicated. This simple observation explains why negative types are much
  simpler and much more appealing in a framework where information is
  guaranteed to be preserved. In previous work that used negative types,
  complicated mechanisms are typically needed to constrain the propagation
  and use of negative values because the surrounding computational framework
  is, generally speaking, careless in its treatment of information.
\item The appeal of negative types can already be gleaned from the credit
  card analogy. The main reason credit card transactions are convenient is
  because they disentangle the propagation of the money from the propagation
  of the services. Not every transaction needs both the resources and
  services to be brought together: it is sufficient to have a promise that
  the money will be paid at some point, as long as the infrastructure can be
  trusted that such promises will be fulfilled. 
\end{itemize}

Similar observations hold for fractional types but we defer these to
Sec.~\ref{sub:frac}. Before proceeding to the main part of the paper, we
briefly summarize the our technical contributions:
\begin{itemize}
\item 
\item 
\item 
\item 
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%
\section{Fractional Types}

%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Summary}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{A Graphical Language for {{langRev}} }

@@

Since the monoidal units cxan be freely introduced and eliminated, in
many diagrams they are omitted and dealt with explicitely when they
are of special interest.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Multiplicative Duality in {{langRev}} }

@@

The usual interpretation of {{b1*b2}} we have both a value of type
{{b1}} and a value of type {{b2}}. This interpretation is maintained
in the precense of fractionals. Hence an {{eta* : 1 <-> inv(b) * b}} is
to be viewed as a fission point for a value of type {{b}} and its
multiplicative inverse {{inv(b)}}. This naturally raises the question --
which value of type {{b}} is produced? The isomorphism {{eta*}} does
not produce any specific value of type {{b}}, instead it produces a
place holder for a value such that the value of type {{inv(b)}} is
always the dual of the chosen one.

@@

\paragraph*{Unification.}
A computational interpretation of this is that {{langRevEE}} acts like
a logic programming system and {{eta*}} acts as the site of creation
of a new typed logic variable of type {{b}} and its corresponding dual
of type {{inv(b)}}. These two values are free to move independantly
through a circuit, but any operations on value affect what operations
are possible on the other since they are both related by an underlying
logic variable. Dually, {{eps*}} acts as a unification site for values
flowing along both branches. If the values are not exactly duals, then
the system fails to unify and hence the program is stuck. If the
values are exactly duals, then unification succeeds and {{eta*}}
returns {{1}}.

@@

This also leads to the exiciting point of view that entangled
particle/anti-particle pairs act as functions in a physical
first-class sense. We can ask the fuction to be applied to a value by
unifying the anti-particle with a particle in a well-known state and
examining is pair. We shall use this point of view in the development
our SAT solver (see Sec. \ref{sec:sat-solver}).

@@

\paragraph*{Annihilation.}
As with {{(+, 0)}} we can construct a {{trace*}} operation. Unlike the
additive {{trace+}}, {{trace*}} gives us the ability to find the
fixpoint of the contraint (expressed with combinators) that is
traced. If the constraint has not satisfying values, there is not
fixpoint possible -- we use the term `annihilation' to describe the
corresponding undefined state of the program. Annihilation and a
non-termination both describe undefined program execution -- while
non-termination characterizes an iteration that fails to terminate,
annihilation characterizes a constraint that has no fixpoint.

\paragraph*{Zero Information.}
As with {{eta+}}, we argue that {{eta*}} constructs no new
information, instead it gives us the ability to explore as many worlds
as permitted by the type {{b}}. So situate ourselves in any specific
world, we have to supply additional information in teh form of
unifying the particle or anti-particle with a particle representing
known information. Hence only by supplying information to a an
entangled pair can we read information from it -- and always exactly
as much information as we supplied.

@@

The construction here is not limited to boolean satisfiability, but
generalises to any constraint satisfaction problem whose search space
can be represented by a recursive type and for which an isomorphic
clone operation can be constructed. That said, we will focus on the
limited case of boolean satisfiability here.

>>>>

In a recent paper~\cite{infeffects}, we argued that, because they include
irreversible physical primitives, conventional abstract models of computation
have inadvertently included some \emph{implicit} computational effects which
we called \emph{information effects}. We then developed a pure reversible
model of computation that is obtained from the type isomorphisms and
categorical structures that underlie models of linear logic and quantum
computing and that treats information as a linear resource that can neither
be erased nor duplicated. In this paper, we show that our pure reversible
model unveils deeper and more elegant symmetries of computation than have
previously been reported. In particular, we expose two notions of duality of
computation: an additive duality and a multiplicative duality that give rise
to negative types and fractional types respectively. Although these types
have previously appeared in the literature (see Sec.~\ref{sec:related}), they
have typically appeared in the context of conventional languages with
information effects, which obscured their appeal and murked their properties.

@@

Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism with a non-trivial and interesting computational
interpretation.  On the left we have a producer that can supply a value of
type $a$ and on the right we have a consumer that demands a value of type
$a$. For the purposes of the example, think of the producer as someone having
\$20.00 and the consumer as someone demanding \$20.00. It is certainly
possible to directly connect the producer and consumer using the algebraic
identity $a=a$: such a connection would be computationally modeled by the
identity function as expected. However, the algebraic proof above implements
this connection in a much more interesting way: the transformation $0 = (-a)
+ a$ is used at the consumer site to create money and a corresponding
debt. The consumer gets the money immediately and the debt travels
``backwards'' towards the producer where it is paid.  In other words, this
computation models something similar to a credit card transaction. We discuss
this example in detail in Sec.~\ref{sub:neg} and only focus on two important
points here:
\begin{itemize}
\item As the example illustrates, negative types corresponds to ``debts.''
  It would clearly be disastrous if debts could be deleted or
  duplicated. This simple observation explains why negative types are much
  simpler and much more appealing in a framework where information is
  guaranteed to be preserved. In previous work that used negative types (see
  Sec.~\ref{sec:related}), complicated mechanisms are typically needed to
  constrain the propagation and use of negative values because the
  surrounding computational framework is, generally speaking, careless in its
  treatment of information.
\item The appeal of negative types can already be gleaned from the credit
  card analogy. The main reason credit card transactions are convenient is
  because they disentangle the propagation of the resources (money) from the
  propagation of the services. Not every transaction needs both the resources
  and services to be brought together: it is sufficient to have a promise
  that the money will be paid at some point, as long as the infrastructure
  can be trusted that such promises will be fulfilled. Computationally, this
  means that negative types allow us to break the dependency between the
  producer and consumer and let each compute independently.
\end{itemize}

Instead of also discussing fractional types and related work in this
introduction, we briefly summarize our main technical contributions and move
on to the heart of the paper where these topics can be explained in their
proper context. Our main contributions are:
\begin{itemize}
\item We extend {{langRev}} our reversible programming language of type
  isomorphisms~\cite{rc2011,infeffects} with a notion of negative types, that
  satisfies the isomorphism $a + (-a) = 0$. The semantics of this extension
  is expressed by having a ``dual'' evaluator that reverses the flow of
  execution for negative values. 
\item We independently extend {{langRev}} with a notion of fractional types,
  that satisfies the isomorphism $a * (1/a) = 1$. The semantics of this
  extension is expressed by introducing logical variables and a unification
  mechanism to model and resolve the constraints introduced by the fractional
  types.
\item We combine the above two extensions into one extension that includes
  both negative and fractional types. The resulting language allows any
  rational number to be used as a type. The types satisfy the same familiar
  and intuitive isomorphisms that are satisfied in the mathematical field of
  rational numbers.
\item We develop programming intuition and argue that negative and fractional
  types ought to be part of the vocabulary of every programmer.
\item We relate our notions of negative and fractional types to previous work
  on continuations. We analyze and relate two apparently unrelated lines of
  work: the first pioneered by
  Filinski~\cite{Filinski:1989:DCI:648332.755574} relates continuations to
  negative types; the second~\cite{Bernardi:2010:CSL:1749618.1749689} relates
  continuations to the fractional types of the Lambek-Grishin calculus.
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{A Graphical Language for {{langRev}} }

@@

Since the monoidal units can be freely introduced and eliminated, in
many diagrams they are omitted and dealt with explicitly when they
are of special interest.

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Fractional Types}

%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Summary}

@@

%%%%%%%%%%%%%%%%%%%%%
\subsection{Multiplicative Duality in {{langRev}} }

@@

The usual interpretation of {{b1*b2}} we have both a value of type
{{b1}} and a value of type {{b2}}. This interpretation is maintained
in the presence of fractionals. Hence an {{eta* : 1 <-> inv(b) * b}} is
to be viewed as a fission point for a value of type {{b}} and its
multiplicative inverse {{inv(b)}}. This naturally raises the question --
which value of type {{b}} is produced? The isomorphism {{eta*}} does
not produce any specific value of type {{b}}, instead it produces a
place holder for a value such that the value of type {{inv(b)}} is
always the dual of the chosen one.

@@

\paragraph*{Unification.}
A computational interpretation of this is that {{langRevEE}} acts like
a logic programming system and {{eta*}} acts as the site of creation
of a new typed logic variable of type {{b}} and its corresponding dual
of type {{inv(b)}}. These two values are free to move independently
through a circuit, but any operations on value affect what operations
are possible on the other since they are both related by an underlying
logic variable. Dually, {{eps*}} acts as a unification site for values
flowing along both branches. If the values are not exactly duals, then
the system fails to unify and hence the program is stuck. If the
values are exactly duals, then unification succeeds and {{eta*}}
returns {{1}}.

@@

This also leads to the exciting point of view that entangled
particle/anti-particle pairs act as functions in a physical
first-class sense. We can ask the function to be applied to a value by
unifying the anti-particle with a particle in a well-known state and
examining is pair. We shall use this point of view in the development
our SAT solver (see Sec. \ref{sec:sat-solver}).

@@

\paragraph*{Annihilation.}
As with {{(+, 0)}} we can construct a {{trace*}} operation. Unlike the
additive {{trace+}}, {{trace*}} gives us the ability to find the
fixpoint of the constraint (expressed with combinators) that is
traced. If the constraint has not satisfying values, there is not
fixpoint possible -- we use the term `annihilation' to describe the
corresponding undefined state of the program. Annihilation and a
non-termination both describe undefined program execution -- while
non-termination characterizes an iteration that fails to terminate,
annihilation characterizes a constraint that has no fixpoint.

\paragraph*{Zero Information.}
As with {{eta+}}, we argue that {{eta*}} constructs no new
information, instead it gives us the ability to explore as many worlds
as permitted by the type {{b}}. So situate ourselves in any specific
world, we have to supply additional information in the form of
unifying the particle or anti-particle with a particle representing
known information. Hence only by supplying information to a an
entangled pair can we read information from it -- and always exactly
as much information as we supplied.

@@

The construction here is not limited to boolean satisfiability, but
generalizes to any constraint satisfaction problem whose search space
can be represented by a recursive type and for which an isomorphic
clone operation can be constructed. That said, we will focus on the
limited case of boolean satisfiability here.
